---
title: "Play Pulse Studios Online Game Data Analysis"
subtitle: "COMP1013"
author: "Jodie Nguyen-Hoang"
date: "`r Sys.Date()`"
output: 
  pdf_document: 
    highlight: zenburn
    number_sections: true
---

\centering

**22158896**

\vspace{6cm}

\raggedright

**Declaration**

By including this statement, we the authors of this work, verify that:

- We hold a copy of this assignment that we can produce if the original is lost or damaged.

- We hereby certify that no part of this assignment/product has been copied from any other student’s work or from any other source except where due acknowledgement is made in the assignment.

- No part of this assignment/product has been written/produced for us by another person except where such collaboration has been authorised by the subject lecturer/tutor concerned.

- We are aware that this work may be reproduced and submitted to plagiarism detection software programs for the purpose of detecting possible plagiarism (which may retain a copy on its database for future plagiarism checking).

- We hereby certify that we have read and understand what the School of Computing, Engineering and Mathematics defines as minor and substantial breaches of misconduct as outlined in the learning guide for this unit.

\newpage

```{r libraries, include = FALSE}
library("tidyverse")
library("tinytex")
library("ggplot2")
library("lubridate")
library("knitr")
library("kableExtra")
library("dplyr")
```

# GitHub

## Version Control

This project uses Git for version control to track changes and manage progress throughout the Play Pulse Studios Online Game Data Analysis. The following commands were used to manage the repository during the project.

### Basic Commands

These commands were used to stage, commit and push files to GitHub.

```{bash, eval = FALSE}
# Check whether files have been changed, added or deleted
git status
# Add all changed files to be committed
git add .
# Save work with a description
git commit -m "Add R Markdown project files"
# Upload files to GitHub for the first time
git push -u origin main
# Upload files to GitHub after the first time
git push
```

### Viewing History

These commands were used to view past commits and track changes throughout the project.

```{bash, eval = FALSE}
# View all past commits
git log
# View past commits in a one line format
git log --oneline
# View past commits with a table or plot
git log --oneline --graph
# To show staged changes that will be committed in the next commit
git diff --staged
```

### Undoing Changes

These commands were used to undo changes when mistakes were made during the analysis.

```{bash, eval = FALSE}
# Undo changes made to a file before it has been committed
# This restores the file back to the last saved version
git checkout -- Analytics-Programming-Project.Rmd
# Undo the last commit but keep the file changes
git reset --soft HEAD~1
# Undo the last commit and discard all file changes
git reset --hard HEAD~1
```

\newpage

```{r load data, cache = TRUE, include = FALSE}
# Assigns players, games, and sessions to their csv file
players = read_csv("players.csv")
games = read_csv("games.csv")
sessions = read_csv("sessions.csv")
```